---
title: "Project"
output:
  bookdown::pdf_document2:
    includes:
      in_header: "longtable.tex"
  html_document: default
date: "2024-10-24"
fig_caption: yes
bibliography: refs.bib
nocite: '@*'
classoption:
  - twocolumn
header-includes:
  - \setlength{\columnsep}{32pt}
---

```{r, include = FALSE}
# 2col format
# title: "Project"
# output:
#   pdf_document:
    # includes:
    #   in_header: "longtable.tex"
#   html_document: default
# date: "2024-10-24"
# classoption:
#   - twocolumn
# header-includes:
#   - \setlength{\columnsep}{32pt}
# knitr::opts_knit$set(kable.force.latex = TRUE)
```


```{r, include=FALSE}
library(tidyverse)
library(knitr)

# Mohamed
library(tidysynth)

# Alejandro
library(maps)
library(mapproj)
library(stargazer)
library(sandwich)
library(lmtest)
library(car)
library(tigris)
library(sf)

knitr::opts_chunk$set(out.width = '100%', out.height = '100%',
                      echo = FALSE, fig.align = 'center', 
                      kable.force.latex = FALSE, message = FALSE)

plot_theme = theme(
  text = element_text(size = 24),
  axis.text = element_text(size = 18)
)
ggplot2::theme_set(plot_theme)
```

## Literature Review
Kendall and Tulip (2018) quantify the effect of zoning on detached house prices in large cities in Australia. They argue that in the absence of zoning, the market for housing is competitive, as there are low barriers to entry and a large number of firms involved in development and construction. Additionally, they note that in US cities with loose zoning laws, prices are close to costs. Thus, the price of a property can be decomposed into the cost of the structure and land, and zoning representing supply distortions. With this approach, they find large price increases attributable to zoning. For example, in 2016, zoning raised the price of detached houses by 73 percent in Sydney and 69 percent in Melbourne. 


## ACS Analysis


```{r,include=FALSE}
acs <- readRDS("acs_sample.rds")
```

This section used the detailed data from the American Community Survey (ACS), an annual survey conducted by the U.S. Census Bureau that captures key social, economic, housing, and demographic information across the United States (https://www.census.gov/programs-surveys/acs/). The ACS dataset is a comprehensive resource for understanding the factors influencing rental housing prices. Specifically, the data includes demographic, geographic, and housing information as well as the locations from which the data was collected. 

Our analysis seeks to identify the variables that significantly impact rental housing prices and quantify their effect. We focus on demographic factors such as household income, geographic indicators like metropolitan status, and structural features of houses to examine their contributions to rental price variations.

Given the self-reporting status of housing value in the dataset, we concluded that rental prices were a more accurate measure of house value. For this reason, we decided to conduct our analysis using this as our dependent variable. By applying robust statistical methods we provide a deep exploration of the determinants of housing rental prices.

### Data Description

The dataset used in this study includes observations on rental housing units from the most recent ACS. Key variables of interest include:

* Monthly Rental Price: The dependent variable, representing the self-reported cost of renting a housing unit.

* Demographic Variables: These include household income, age of the householder, population density.

* Housing Characteristics: Information about the number of bedrooms, age of the property, kitchen and plumbing inclusion, and room number.

* Geographic Variables: Metropolitan versus rural location, and proximity to major urban centers.

The data was cleaned to ensure accuracy and consistency. The total number of observations was 56,835,321 with 29 features but after removing rows with missing or NA values for rental price or explanatory variables (76.5%) and selecting important features, we were left with 13,341,750 observations with 15 features. 

We are not including taxincl and insincl because those columns only have zeros. Likewise, we are excluding debt_incurred since it only has FALSE. Finally, year_from_start was removed given it was perfectly correlated with year.

```{r}
# Not including taxincl, insincl and debt_incurred because first two only have zero
# and last one only has false

#year_from_start is perfectly correlated with year so we removed it.

data <- acs %>% select(year, density, metro, fips,
                        hhincome, puma,
                        rooms, builtyr2, unitsstr, bedrooms,
                        age, married, male,
                        complete_plumbing, has_kitchen, statefip,
                       cpi_rent) %>%
  na.omit() %>%
  mutate(year = as.factor(year))
  
```

The dataset contains observations from counties across the United States. However, due to challenges in obtaining complete information and the presence of missing values in the FIPS column, it does not provide a clear representation of individual counties. Instead, we aggregated the data at the state level. The map below highlights the states, with darker shades indicating a higher number of observations in the dataset.


```{r}
# Load the US states map data
us_states <- map_data("state")

# FIPS to state name mapping
fips_to_state <- data.frame(
  statefip = c(01, 02, 04, 05, 06, 08, 09, 10, 11, 12, 13, 15, 16, 17, 18, 19, 20, 21, 22, 23,
           24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41,
           42, 44, 45, 46, 47, 48, 49, 50, 51, 53, 54, 55, 56),
  state = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado",
            "connecticut", "delaware", "district of columbia", "florida", "georgia",
            "hawaii", "idaho", "illinois", "indiana", "iowa", "kansas", "kentucky",
            "louisiana", "maine", "maryland", "massachusetts", "michigan",
            "minnesota", "mississippi", "missouri", "montana", "nebraska", "nevada",
            "new hampshire", "new jersey", "new mexico", "new york", "north carolina",
            "north dakota", "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island",
            "south carolina", "south dakota", "tennessee", "texas", "utah", "vermont",
            "virginia", "washington", "west virginia", "wisconsin", "wyoming")
)

# Join the datasets to convert FIPS to state names
data <- data %>%
  left_join(fips_to_state, by = "statefip")


# Count the number of times each state appears in the dataset
state_counts <- data %>%
  count(state, name = "frequency")

# Join the state counts with the map data
us_states_data <- us_states %>%
  mutate(region = as.character(region)) %>%
  left_join(state_counts, by = c("region" = "state"))

# Create the map
ggplot(us_states_data, aes(long, lat, group = group, fill = frequency)) +
  geom_polygon(color = "white", size = 0.2) +
  scale_fill_gradient(
    low = "lightblue", 
    high = "darkblue", 
    na.value = "lightgray", 
    name = "Frequency"
  ) +
  theme_void() +
  labs(
    title = "US States Shaded by Frequency in Dataset",
    caption = "Data source: ACS"
  )

```

The following statistics provide an overview of the data distribution, highlighting significant variation in rental prices, income levels, and property characteristics. Such variation underscores the need for robust multivariate analysis to disentangle the effects of different predictors. 

```{r}
# Select only numeric variables
numeric_data <- data %>% select_if(is.numeric)

# Summarize the statistics for numeric columns
summary_table <- data.frame(
  Variable = colnames(numeric_data),
  Mean = round(sapply(numeric_data, mean, na.rm = TRUE), 2),
  Variance = round(sapply(numeric_data, var, na.rm = TRUE), 2),
  Min = round(sapply(numeric_data, min, na.rm = TRUE), 2),
  Max = round(sapply(numeric_data, max, na.rm = TRUE), 2)
)

rownames(summary_table) <- 1:nrow(summary_table)

# Print the table nicely
kable(summary_table, format = "markdown", caption = "Summary Statistics for Numeric Variables", longtable = F)
```

One variable that caught our attention was age. At first, we were surprised and confused by the presence of ages such as 0 years old or very young ages. However, after further investigation, we discovered that the ACS collects data on every person living in each household to ensure a more comprehensive and accurate measure. As a result, the age variable includes values ranging from 0 to 97.

The following summary of the factor features provides further insight into our dataset. As indicated by the FIPS feature, we have data from every state in the country, as shown by the 51 unique values in this feature. California (represented by FIPS code 06) is the most common state in the sample dataset. Additionally, the "year" feature contains 18 levels, corresponding to 18 years of data in the sample, with 2014 being the most commonly represented year.

```{r}
# Select only factor (categorical) variables
factor_data <- data %>% select(-fips) %>%
  select_if(is.factor)

# Summarize the statistics for factor columns
summary_table2 <- data.frame(
  Variable = colnames(factor_data),
  `Number of Levels` = sapply(factor_data, function(col) length(unique(col))),
  `Most Frequent Level` = sapply(factor_data, function(col) names(sort(table(col), decreasing = TRUE)[1])),
  `Frequency of Most Frequent Level` = sapply(factor_data, function(col) max(table(col)))
)

# Print the table nicely
kable(summary_table2, format = "markdown", caption = "Summary Statistics for Factor Variables", longtable = F)
```


### Methodology

To analyze the determinants of rental prices, we adopted a multivariate linear regression model. This method allows us to evaluate the individual contribution of each explanatory variable while controlling for the effects of others. The steps taken in the analysis are as follows:

Variable Selection: Using domain knowledge and exploratory data analysis, we identified variables with potential relevance to rental price determination.

Model Specification:

The primary model is specified as:

$$
Y = \beta_0 + \gamma_i * year_i + \theta_j * state_j + \beta_k * X_k + \epsilon
$$

Where:

* $Y$ represents the dependent variable (Rent price).
* $X_1, X_2, X_3,\dots, X_n$ are the independent variables (e.g., [density, metro, hhincome, rooms, builtyr2, unitsstr, bedrooms, age, married, male, complete_plumbing, has_kitchen]).
* $year_i$ is the dummy variable for the year with $i \in \{1,\dots, 18\}$ as we're controlling for 18 years of data.
* $state_j$ is the dummy variable for the state with $j \in \{1,\dots, 51\}$ as we're controlling for all 50 states and the District of Columbia.
* $\beta_0$ is the intercept of the regression model.
* $\beta_1, \beta_2, \beta_3,\dots, \beta_n$ are the coefficients that represent the relationship between each independent variable and the dependent variable.
* $\gamma_i$ are the coefficients that represent the relationship between each year and the dependent variable.
* $\theta_j$ are the coefficients that represent the relationship between each state and the dependent variable.
* $\epsilon$ is the error term, capturing unobserved factors.

Model Estimation: To estimate the coefficients, we employed Ordinary Least Squares (OLS) regression. Diagnostic tests were conducted to validate the model assumptions:

Multicollinearity: Variance Inflation Factors (VIF) were calculated to ensure no strong multicollinearity among the predictors.
The model was implemented using the following code, which provided the estimated coefficients:

```{r, results = 'asis'}
data <- data %>% select(-fips)
fit.lm_acs <- lm(cpi_rent ~ ., data)
fit.coeftest = coeftest(fit.lm_acs, vcov=vcovHC, type="HC0")
stargazer(fit.coeftest, omit = c("year", "state"))
```

To ensure robustness, we applied robust standard errors as displayed below:

The VIF scores were computed to further validate the model:

```{r}
#vif <- vif(fit.lm_acs)
print(vif)
```

Before analyzing the results, we note that all estimates are statistically significant. After applying robust standard errors, each estimated coefficient remains significant. Furthermore, we computed the Variance Inflation Factor (VIF) scores to ensure the validity of the model. All VIF scores are below 5, indicating no strong multicollinearity. This confirms that our model is both reliable and robust.

Model Overview:
The regression model included a total of 79 predictors and was fitted using a dataset with 13,341,750 observations. The R-squared of 0.4267 indicates that 42.67% of the variance in CPI Rent is explained by the included predictors. The F-statistic ($ 1.257e^5, p < 2.2e^{-16}$) demonstrates that the overall model is statistically significant.

Key Results:

Time Trends:

Yearly dummy variables were included to capture temporal trends. Statistically significant coefficients ($p < 0.001$) indicate clear yearly variation in CPI Rent. For example, CPI Rent increased significantly in 2020 compared to the base year($\hat{\gamma_{2020}} = 49.41$, $p < 2e^{-16}$) and declined in 2011 compared to the base year ($\hat{\gamma_{2011}} = -6.86$, $p < 2e^{-16}$).

Population Density:

The coefficient for population density was small but highly significant ($\hat{\beta_{density}} = 0.003$, $p < 2e^{-16}$) indicating that higher population density is associated with marginal increases in the Consumer Price Index (CPI) Rent. However, given that population density values across most regions in the dataset are not significantly high, this suggests that density has a notable impact primarily in areas with exceptionally high population densities. This means the effect of population density on CPI Rent is significant, but only for a select group of regions where density levels are substantially higher than average.

Metro Areas:
The presence of metro designation (`metro == TRUE`) substantially increased CPI Rent ($\hat{\beta_{metro}} = 166.5$,$p < 2e^{16}$).

State Effects:
State fixed effects (`fips_state`) showed considerable variation. For example, being in state 15 (Hawaii) significantly increased CPI Rent ($\hat{\theta_{15}} = 527.3$, $p < 2e^{-16}$) relative to the baseline.

Housing Characteristics:
Variables like the number of rooms ($\hat{\beta_{rooms}} = 22.44$, $p < 2e^{-16}$) and the presence of complete plumbing ($\hat{\beta_{plumbing}} = 125.0$, $p < 2e^{-16}$) were positively associated with CPI Rent. Conversely, older buildings (`builtyr2`) and lack of a kitchen had significant negative effects.

3. Variance Inflation Factor (VIF) Analysis:
The VIF values for all predictors were below 3, except for rooms (GVIF =  $3.039$), indicating no severe multicollinearity. This strengthens the robustness of the model's coefficient estimates.

4. Residual Analysis:
The residual standard error was $340.4$, indicating the average deviation between observed and predicted values. Residuals ranged from $-6,398.4$ to $4,859.9$, with a median near zero, suggesting an overall balanced fit.

5. Limitations and Future Research:
Although the model captures significant variance, $57.33%$ of CPI Rent variability remains unexplained. Including additional predictors like economic trends, housing policies, and neighborhood amenities could enhance model performance.

## Data and Description
For rents, we use the data set of Fair Market Rents provided by the U.S. Department of Housing and Urban Development, which provides annual data on the 40th percentile of rent on a county level from 1995 onwards. Data on sales prices come from the Survey of Construction, administered by the U.S. Census Bureau and U.S. Department of Housing and Urban Development, and come in the form of quarterly median sales price of houses sold in the United States, from the start of 1963 until July 2024. Construction price indices for houses sold and houses under construction are also sourced from the Survey of Construction data. Indices for houses under construction are monthly from 1963, while indices for houses sold are quarterly from 1963. Both are Laspreyes price indices, with the houses sold index being based on the cost of building the house and the value of the land, and the houses under construction index being based on only the cost of building the house. The CPI series used is the index for all urban consumers and all items, and comes from the Bureau of Labor Statistics, and is available on a monthly basis until September 2024.

(Realtor.com needs to be described, Max should already be doing this)

### Analyzing Supply Costs

```{r}
constr = read_csv("CausalAnalysis/Final/ConstructionPriceIndex/CONSTR.csv")

plotConstrVars = function(vars, after = 2012){
  constr2 = constr %>% 
    filter(year >= after) %>% 
    pivot_longer(all_of(vars))
  
  breaks = (constr2$date %>% unique)[seq(1, length(constr$date), by = 6)]
  
  ggplot(constr2) +
    geom_line(aes(x = date, y = value)) +
    facet_wrap(name ~ ., ncol = 1, scales = "free_y") +
    scale_x_continuous(breaks = breaks,
                       guide = guide_axis(angle = 45))
}
```


Traditional economic models state that the price of a good is affected by the demand and supply functions. A change in the supply function for housing from, for example, an increase in construction costs or the price of land would increase the price of housing. We can control for these factors with the Survey of Construction's price indices. We divide median price of purchased houses by the houses sold index to yield the variable sold_corrected: the price of houses relative to cost of construction and land, thus controlling for the cost of supply.

House price data shows a significant break around 2011, in the aftermath of the Great Recession. Given the massive effect the Great Recession had on the housing market and the large government response, the sharp increase in housing prices may represent a structural break in the data. For this reason we prefer to limit our data range to 2012 onwards.
```{r chunkname, fig.cap = 'Map'}
# Set tigris options for retrieving data
options(tigris_use_cache = TRUE)

# Set tigris options
options(tigris_use_cache = TRUE)

# Fetch PUMA shapefiles for the required states
# Use a loop or lapply to fetch PUMAs for all statefips in your dataset
puma_shapefiles <- lapply(unique(data$statefip), function(state) {
  pumas(state = state, cb = TRUE, year = 2020) # Download PUMAs for the given state
})

# Combine all state shapefiles into a single sf object
puma_combined <- do.call(rbind, puma_shapefiles)

# Filter PUMAs based on the dataset
highlighted_pumas <- puma_combined %>%
  filter(as.integer(STATEFP20) %in% data$statefip & as.integer(PUMACE20) %in% data$puma)

map_plot <- ggplot() +
  geom_sf(data = puma_combined, fill = "lightgray", color = "white", size = 0.2) + # Background PUMAs
  geom_sf(data = highlighted_pumas, fill = "red", color = "black", size = 0.4) +  # Highlighted PUMAs
  coord_sf(
    xlim = c(-130, -60), # Longitude limits for continental U.S.
    ylim = c(20, 55)     # Latitude limits for continental U.S.
  ) +
  theme_minimal() +
  labs(title = "Highlighted PUMAs",
       caption = "Source: U.S. Census Bureau") +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 12),
    axis.text = element_blank(),  # Remove axis labels for clarity
    axis.ticks = element_blank()
  )

map_plot
```

```{r figma, fig.cap = 'Graph of sold corrected 1992-2023'}
plotConstrVars("sold_corrected", 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0))
```
See Figure \@ref(fig:figma)
We contrast prices controlled for inflation only with CPI (cpi_mspus) to prices controlled for with the houses sold index. The most striking feature of this figure is that while cpi_mspus sees a sharp increase in prices starting at roughly 2020, sold_corrected remains stable. Additionally, while cpi_mspus increases steadily until roughly 2018 and stabilizes until 2020, sold_corrected starts stable and then decreases.
```{r, message=FALSE, fig.cap = 'Comparison between cpi mspus and sold corrected'}
plotConstrVars(c("sold_corrected", "cpi_mspus"))
```
While the methodology does not imply a causal interpretation, it seems highly likely that trends in the housing price are caused by trends in construction costs. Supply costs are a major factor in the cost of almost all goods, and it is hard to conceive of any confounding variable correlated with price and supply cost that could dwarf the effect of supply. Additionally, we graph inflation-controlled housing prices (cpi_mspus) and the inflation-controlled houses sold index (cpi_sold) to show that trends in both coincide.
```{r}
plotConstrVars(c("cpi_mspus", "cpi_sold"))
```

One factor that could affect house sales price on the demand side is the interest rate: a lower interest rate increases the demand for housing due to more favorable terms on mortgages. To eliminate this factor, we also use the price of houses purchased with cash (full price up front with no mortgage or loan) which should be unaffected by the interest rate. After controlling with the houses sold index, we see the same general trend, albeit with much more volatility. The average price is stable until 2018, starts decreasing until 2020, then seems to stabilize again.
```{r}
plotConstrVars("sold_corrected_cash")
```

Additionally, we can control using the houses under construction index instead, which only accounts for construction costs, and contrast with the price controlled by houses sold index. Both variables appear near-identical, indicating construction costs as the main driver of the effect we see.
```{r}
plotConstrVars(c("constr_corrected", "sold_corrected"))
```

### Zoning synthetic controls
The practice of zoning, or restricting what types of construction can occur on certain areas, is often blamed for driving up the price of housing by artificially limiting supply. One particular form of zoning that receives significant attention is single-family zoning, only allowing the construction of single-family homes. As part of the comprehensive Minneapolis 2040 plan, taking effect January 1, 2020, Minneapolis passed a law ending single-family zoning, a feat few other cities have matched. Additionally, the city implemented inclusionary zoning, requiring developers to have some of their units to be rented at prices below certain percentages of median income in the area.

To assess the effectiveness of these policies in reducing local rents, we utilize a synthetic control approach, as we have no reason to believe the parallel trends assumption should hold. The control group is comprised of metro areas as designated by the U.S. Department of Housing and Urban Development in the states of Minnesota (which contains Minneapolis) and Wisconsin. One limitation of our data is that the metro area containing Minneapolis also contains the slightly smaller neighboring city of St. Paul, which was not directly affected by the laws passed. This should dilute the overall effect, but so long as there are no simultaneous shocks to housing prices in St. Paul, a causal interpretation is still possible. We could also expect to see spillover effects decreasing rent in St. Paul as landlords compete with lower prices. For predictive variables, we use population in 2010 and market factors from 2016.

Our results indicate a increase in rent in the metro area containing Minneapolis and St. Paul compared to the synthetic control. As outlined in Abadie et al. (2010), we compute Fisher's Exact p-value as 0.5714, indicating no signifiant change in rents.
```{r}
fmr_synth = read_csv("CausalAnalysis/Final/Minnesota/FMR_synth.csv")

treat_date = 2020
treated_unit = "Minneapolis-St. Paul-Bloomington, MN-WI HUD Metro FMR Area"

synth = fmr_synth %>%
  synthetic_control(outcome = price,
                    unit = areaname22,
                    time = year,
                    i_unit = treated_unit,
                    i_time = treat_date) %>% 
  generate_predictor(time_window = 2016,
                     populat = pop2010,
                     mdom = median_days_on_market,
                     nlc = new_listing_count,
                     plc = pending_listing_count,
                     pic = price_increased_count,
                     msf = median_square_feet
                     ) %>%
  generate_weights() %>% 
  generate_control()

plot_trends(synth) + plot_theme
plot_differences(synth) + plot_theme
```

\newpage

## References


